---
title: "Basic Concepts of R"
author: "MD Tanvir Anjum"
date: "6/11/2020"
output: html_document
---

<h1>Data Types- R Objects and Attributes</h1> 

<h2>Objects</h2>

R has five basic or "atomic" classes of objects: 

  * Character
  * numeric (real numbers)
  * integer 
  * complex
  * logical (True/False)
  
The most basic object is a vector 
 
 * A vector can only contain  objects of the same class
 * But: The one exception is a <em>list</em>, which represented as a vector but can contain objects of different classes 
 (indeed, that's usually why we use them)

Empty vectors can be created with the vector() function.

<h2>Numbers</h2>
  
  * Numbers in R a generally treated as numeric objects (i.e. double precision real numbers)
  * If you explicitly want an integer, you need to specify the L suffix 
```{r echo=TRUE, message=FALSE}
x <- 1 #this will output a numeric value 
class(x) 
#if you want to create integer explicitly 
y <- 1L
class(y) 
```
  
  or you can forcefully change the value to integer by: 
```{r echo=TRUE, message=FALSE}
x <- 1
class(x)
y <- as.integer(x)
class(y)
```

  * There are also a special number Inf which represents infinity; e.g. 1/0; Inf can be used in ordinary calculations
  
```{r}
x <- 1/0 
x
y <- 1/Inf
y
```
  
  * The value NaN represents an undefined value ("not a number"); e.g. 0/0; NaN can also be thought as missing value 
```{r}
x <- 0/0
x
```
  
<h2>Attributes</h2>
R objects can have attributes 

```{r echo=FALSE, paged.print=TRUE}
library(knitr)
library(kableExtra)
text_tbl <- data.frame(
  Attribute = c("class", "comment", "dim", "dimnames", "names", "row.names", "tsp", "levels"),
  Description = c(
    "The class of the object.",
    "A comment on the object; often a description of what the object means. ", 
    "Dimensions of the object.",
    "Names associated with each dimension of the object.",
    "Returns the names attribute of an object. Results depend on object type; for example, it returns the name of each data column in a data frame or each named object in an array.",
    "The name of each row in an object (related to dimnames).",
    "Start time for an object. Useful for time series data.",
    "Levels of a factor."
  )
)

kable(text_tbl) %>%
  kable_styling(full_width = F) %>%
  column_spec(1, bold = T, border_right = T) %>%
  column_spec(2, width = "30em", background = "white")
```

As an example:

```{r}
x <- cbind(a=1:3, pi=pi) # simple matrix w/ dimnames
attributes(x)

```


